---
title: "Model Extension"
author: "Sadie Cwikiel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Model Set-Up

```{r}
## Parameters
# parrotfish growth rate
s <- 0.49 # intrinsic rate of growth s of parrotfish 

# rate macroalgae directly overgrow coral per year
alpha <- 0.1 # coral overgrown by macroalgae at a rate a

# rate macroalgae spread vegetatively over algal turfs per year
gamma <- 0.8 # at a rate γ (gamma)

# rate of coral recruitment to algal turfs per year; coral always overgrow algal turfs
r <- 1 # corals recruit to and overgrow algal turfs at a rate r

# coral mortality -- natural coral mortality accounts for 2-4% per year, predation for 30%, and remaining mortality from coral disease or tissue damage following bleaching
d <- 0.44 # natural mortality rate of d

# grazing rate g can change, right now im setting it between the critical values 0.18 and 0.39
g <- 0.4 # set grazing rate 

# maximum parrotfish carrying capacity
#beta <- C

# fishing pressure 
f <- 0
```


```{r}
# storing colors 
Mcol <- 'limegreen' # plot color for macropalgae 
Tcol <- 'darkolivegreen4' # plot color for turf algae 
Ccol <- 'chocolate1' # plot color for coral 
Pcol <- 'darkturquoise' #plot color for parrotfish population
```

```{r}
# setting time and inital values 
# time set 0 to 100
tset <- seq(from = 0, to = 100, length.out = 20000) #set time from 0 to 100 

# set initial condition for the parameters, I just set these randomly to be equal to 1 as the paper says
M0 <- 0.5
C0  <- .3
T0  <- 1 - M0 - C0 
P0 <- 1
```

```{r}
### Temperature Dependence Variables and Temperature Data ###
# create temperature vectors
# min = 26 deg C, max = 33 deg C, same length out as the tset

# bleaching threshold
bleaching_temp <- 30

# random temp vector 
### 100 days, random temperatures between 26 and 31 degrees C
random_temp <- runif(n=100, min=26, max=32)

# heatwave vector
heatwave <- runif(n=10, min = 30, max = 32)

# non heatwave 

# degree heating day/week?
# dhd <- if random_temp[i-1] > 30 , -- should calculate the difference between the temps? 




```

#### TEMP SET UP QUESTIONS:
- how long should the temp vector be? tset goes from 0-100, so should the temp vector be 100 days? should it be a whole year? 
- i've created a random temp set, but how to create thermal regimes that i want to test? like going above the threshold once and then letting the system go back to normal?


# Modified Simulations 1, 2, and 3 with Temperature Dependence

## Simulation 1: Original Model Without Parrotfish


```{r}
### Simulation 1: M, C, T dynamics in the absence of parrotfish grazing ###

# creating holding vectors and filling initial conditions
M.simu <- NaN*tset 
M.simu[1] <- M0
T.simu <- NaN*tset
T.simu[1] <- T0
C.simu <- NaN*tset
C.simu[1] <- C0
P.simu <- NaN*tset
P.simu[1] <- P0 #start P squiggle (only using P squiggle, so P squiggle = P; g = P squiggle because g(P) = P/beta, and we assume beta = 1 (max carrying capacity, which is the same as max coral cover), so setting initial g value to P0

#for loop
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu[i-1]
	C <- C.simu[i-1]
	Tt <- T.simu[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simu[i] <- M.simu[i-1] + dM
	C.simu[i] <- C.simu[i-1] + dC
	T.simu[i] <- 1 - C.simu[i] - M.simu[i]
}
```


## Simulation 2: habitat as primary limiting resource

```{r}
####  Simulation 2: WITH PARROTFISH -- habitat as primary limiting resource ####

## Notes and Assumptions ##
# parrotfish logistic growth changes the grazing rate (g = P, see model assumptions) and therefore impacts M, C, and T dynamics
# assume that f = 0 (no fishing)
# g(P) = alpha*P/beta where alpha is a positive constant. # alpha = g_max, assumed to be 1, so g(P) = P/beta, which guarantees that the grazing intensity. arrives at a maximum only if f = 0 and P(t) reaches maximum carrying capacity, requiring no limitation from habitat (i.e., K(C) = 1)
# so we can assume for this simulation that beta = K(C) = C
# P = P squiggle here, which is population size relative to maximum carrying capacity (beta)
# g(P) = P squiggle = P, so in the equation for dM, we've replaced g with P

# creating holding vectors and filling initial conditions
M.simu2 <- NaN*tset 
M.simu2[1] <- M0
T.simu2 <- NaN*tset
T.simu2[1] <- T0
C.simu2 <- NaN*tset
C.simu2[1] <- C0
P.simu2 <- NaN*tset
P.simu2[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu2[i-1]
	C <- C.simu2[i-1]
	Tt <- T.simu2[i-1]
	P <- P.simu2[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	
	# calculating total population size
	M.simu2[i] <- M.simu2[i-1] + dM
	C.simu2[i] <- C.simu2[i-1] + dC
	T.simu2[i] <- 1 - C.simu2[i] - M.simu2[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu2[i] <- P.simu2[i-1] + dP
}
```


## Simulation 3: food as primary limiting resource
```{r}
####  Simulation 3: WITH PARROTFISH -- food availability as primary limiting resource ####

# K(C) = 1 - zC where 0 <= z < 1
# in supplemental materials, they start with z = 0.5, but they also use z = 0.1, 0.3, 0.6, and 0.9. They find that hysteresis is possible for each value of z.

#set z value
z <- 0.5

# creating holding vectors and filling initial conditions
M.simu3 <- NaN*tset 
M.simu3[1] <- M0
T.simu3 <- NaN*tset
T.simu3[1] <- T0
C.simu3 <- NaN*tset
C.simu3[1] <- C0
P.simu3 <- NaN*tset
P.simu3[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu3[i-1]
	C <- C.simu3[i-1]
	Tt <- T.simu3[i-1]
	P <- P.simu3[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity K(C) = 1 - zC
	dP <- ((s*P)*(1 - (P/(1-(z*C))) - f*P)) * dt
	
	# calculating total population size
	M.simu3[i] <- M.simu3[i-1] + dM
	C.simu3[i] <- C.simu3[i-1] + dC
	T.simu3[i] <- 1 - C.simu3[i] - M.simu3[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu3[i] <- P.simu3[i-1] + dP
}
```

