---
title: "EEMB 179 Final Project Code"
author: "Sadie Cwikiel"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

set up equations 
$$
\begin{align}
\frac{d M}{dt}  &= \alpha M C - \frac{ gM}{(M + T)} + \gamma M T\\ 
\newline
\frac{d C}{dt} &= rTC - dC - \alpha M C\\
\newline
\frac{d T}{dt} &= \frac{ gM}{(M + T)} - \gamma M T + dC\\
\newline
\frac{d P}{dt} &= s P(1 - \frac{P}{\beta K(C)}) - fP\\
\newline
\frac{d \tilde{P}}{dt} &= s \tilde{P}(1 - \frac{\tilde{P}}{K(C)}) - f\tilde{P}\\
\end{align}
$$

Equations 
```{r}
### Macroalgae ###
# dMdt <- (alpha*M*C) - ((g*M)/(M + T)) + (gamma*M*T)


### Coral ###
# dCdt <- (r*T*C) - (d*C) - (alpha*M*C)


### Turf algae ###
# dTdt <- (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)
```

Parameters 
```{r}
# parrotfish growth rate
s <- 0.49 # intrinsic rate of growth s of parrotfish 

# rate macroalgae directly overgrow coral per year
alpha <- 0.1 # coral overgrown by macroalgae at a rate a

# rate macroalgae spread vegetatively over algal turfs per year
gamma <- 0.8 # at a rate γ (gamma)

# rate of coral recruitment to algal turfs per year; coral always overgrow algal turfs
r <- 1 # corals recruit to and overgrow algal turfs at a rate r

# coral mortality -- natural coral mortality accounts for 2-4% per year, predation for 30%, and remaining mortality from coral disease or tissue damage following bleaching
d <- 0.44 # natural mortality rate of d

# grazing rate g can change, right now im setting it between the critical values 0.18 and 0.39
g <- 0.4 # set grazing rate 

# maximum parrotfish carrying capacity
#beta <- C

# fishing pressure 
f <- 0

```
 
Variables 
```{r}
###  Variables ###

#M <- macroalgae 
  # M <- 1 - C - T
#C <- coral 
  # C <- 1 - M - T
#T <- turf algae 
  # T <- 1 - M - C

# M + C + T = 1 particular region of the seabed is covered entirely by macroalgae (M), coral (C), and algal turfs (T) (variables)

# g # parrotfish graze macroalgae without distinction from algal turfs at a rate g


# storing colors 
Mcol <- 'limegreen' # plot color for macropalgae 
Tcol <- 'darkolivegreen4' # plot color for turf algae 
Ccol <- 'chocolate1' # plot color for coral 
Pcol <- 'darkturquoise' #plot color for parrotfish population
```

##  Simulation 1: Original Model (no parrotfish)

```{r}
# setting time and inital values 
# time set 0 to 100
tset <- seq(from = 0, to = 100, length.out = 20000) #set time from 0 to 100 

# set initial condition for the parameters, I just set these randomly to be equal to 1 as the paper says
M0 <- 0.5
C0  <- .3
T0  <- 1 - M0 - C0 
P0 <- 1
```

```{r}
### Simulation 1: M, C, T dynamics in the absence of parrotfish grazing ###

# creating holding vectors and filling initial conditions
M.simu <- NaN*tset 
M.simu[1] <- M0
T.simu <- NaN*tset
T.simu[1] <- T0
C.simu <- NaN*tset
C.simu[1] <- C0
P.simu <- NaN*tset
P.simu[1] <- P0 #start P squiggle (only using P squiggle, so P squiggle = P; g = P squiggle because g(P) = P/beta, and we assume beta = 1 (max carrying capacity, which is the same as max coral cover), so setting initial g value to P0

#for loop
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu[i-1]
	C <- C.simu[i-1]
	Tt <- T.simu[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simu[i] <- M.simu[i-1] + dM
	C.simu[i] <- C.simu[i-1] + dC
	T.simu[i] <- 1 - C.simu[i] - M.simu[i]
}
```


```{r}
# plot for M/C/T dynamics in the absence of parrotfish grazing

# plot macroalgae
plot(x = tset, y = M.simu,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu,
      lwd = 2, col = Ccol)

title("M, C, and T Dynamics Without Parrotfish")

legend(x = 60, y = 1,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```
This figure shows how macroalgae, turf algae, and coral cover change as a function of time in the absence of grazing by parrotfish. As macroalgae decreases, coral has more room to grow, so coral cover increases. 


##  Simulation 2: WITH PARROTFISH -- habitat as primary limiting resource

Assumptions:
- Parrotfish populations grow logistically and influence the grazing rate, therefore impacting M, C, and T dynamics.
- Parrotfish rely on coral for habitat, and parrotfish population growth in simulation 2 is limited by coral cover.
- First, we assume that fishing pressure (f) = 0.
- The grazing rate of parrotfish is described as g(P) = (alpha*P)/beta. Alpha is a positive constant and is equal to the maximum grazing rate, so is assumed to be 1. Therefore, g(P) = P/beta, which guarantees that the grazing intensity arrives at a maximum only if f = 0 and P(t) reaches maximum carrying capacity (beta), requiring no limitation from habitat (i.d., K(C) = 1).
- For this simulation, we assume that beta = K(C) = C. 
- P squiggle is the equation of parrotfish population growth with respect to beta (max carrying capacity), and we use P to signify P squiggle in the simulation below. 
- P (squiggle) has been substituted in for g in the macroalgae cover equation because g(P) = Psquiggle. 

```{r}
####  Simulation 2: WITH PARROTFISH -- habitat as primary limiting resource ####

## Notes and Assumptions ##
# parrotfish logistic growth changes the grazing rate (g = P, see model assumptions) and therefore impacts M, C, and T dynamics
# assume that f = 0 (no fishing)
# g(P) = alpha*P/beta where alpha is a positive constant. # alpha = g_max, assumed to be 1, so g(P) = P/beta, which guarantees that the grazing intensity. arrives at a maximum only if f = 0 and P(t) reaches maximum carrying capacity, requiring no limitation from habitat (i.e., K(C) = 1)
# so we can assume for this simulation that beta = K(C) = C
# P = P squiggle here, which is population size relative to maximum carrying capacity (beta)
# g(P) = P squiggle = P, so in the equation for dM, we've replaced g with P

# creating holding vectors and filling initial conditions
M.simu2 <- NaN*tset 
M.simu2[1] <- M0
T.simu2 <- NaN*tset
T.simu2[1] <- T0
C.simu2 <- NaN*tset
C.simu2[1] <- C0
P.simu2 <- NaN*tset
P.simu2[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu2[i-1]
	C <- C.simu2[i-1]
	Tt <- T.simu2[i-1]
	P <- P.simu2[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	
	# calculating total population size
	M.simu2[i] <- M.simu2[i-1] + dM
	C.simu2[i] <- C.simu2[i-1] + dC
	T.simu2[i] <- 1 - C.simu2[i] - M.simu2[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu2[i] <- P.simu2[i-1] + dP
}

```

```{r}
# plot macroalgae
plot(x = tset, y = M.simu2,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu2,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu2,
      lwd = 2, col = Ccol)

title("M, C, and T Dynamics with Parrotfish Grazing \n (Habitat as Primary Limiting Resource)")

legend(x = 60, y = 1,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```
This figure shows how macroalgae, turf algae, and coral cover change with time if parrotfish grazing is present and fishing effort is 0. The final equilibrium state is the same as the scenario without parrotfish.

```{r}
# plot macroalgae
plot(x = tset, y = P.simu2,
     type = 'l', las = 1, lwd = 2, col = Pcol,
     xlab = 'Time', ylab = 'Population Size Relative to Maximum Carrying Capacity')

title("Parrotfish Population Growth \n (Habitat as Primary Limiting Resource)")

```
Parrotfish abundance relative to maximum carrying capacity (beta) over time. The population declines rapidly at the beginning of the simulation when coral cover is low because the proportion of coral cover is assumed to be the same as the carrying capacity in this habitat-limited scenario. As coral cover increases and levels off throughout the simulation, the parrotfish population does the same. 

```{r}
### Bifurcation diagram of fishing effort parameter vs. coral cover equilibrium value

# sequence of values for f (fishing pressure)
fset <- seq(from = 0, to = 1,length.out = 50)

# holding vectors for equilibrium values of C, M, and T
Cstarset <- NaN*fset 
Mstarset <- NaN*fset
Tstarset <- NaN*fset
Pstarset <- NaN*fset

# for loop for fishing pressure bifurcation diagram
for(j in 1:length(fset)){
  # assign the value of Y
	f <- fset[j] 
	
	# create a holding vector for predator and prey populations
	# and fill with initial conditions
	C.simu_f <- NaN*tset
	C.simu_f[1] <- 1   
	M.simu_f <- NaN*tset 
	M.simu_f[1] <- 1
	T.simu_f <- NaN*tset 
	T.simu_f[1] <- 1
	P.simu_f <- NaN*tset 
	P.simu_f[1] <- 1
	
	for(i in 2:length(tset)){
	  # calculating change in time
		  # calculating change in time
	  dt <- tset[i]-tset[i-1] 
	
	  # set variable equal to the simulation value at the previous timestep
	  M <- M.simu_f[i-1]
	  C <- C.simu_f[i-1]
	  Tt <- T.simu_f[i-1]
	  P <- P.simu_f[i-1]
		
  	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
  	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	  # calculating change in population size -- replace g with P squiggle (P)
  	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
  	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
  	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt

	
  	# calculating total population size
  	M.simu_f[i] <- M.simu_f[i-1] + dM
  	C.simu_f[i] <- C.simu_f[i-1] + dC
  	T.simu_f[i] <- 1 - C.simu_f[i] - M.simu_f[i]
	  # P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to   coral cover.
  	P.simu_f[i] <- P.simu_f[i-1] + dP
  	} #end of the inner for loop
	
	# storing last population size (equilibrium population size) in holding vectors for equilibrium population sizes
	Cstarset[j] <- C.simu_f[length(tset)]
	Mstarset[j] <- M.simu_f[length(tset)]
	Tstarset[j] <- T.simu_f[length(tset)]
	Pstarset[j] <- P.simu_f[length(tset)]
	
} #end of outer for loop that iterates over each element in Yset

```

```{r}
## Plot for fishing pressure vs. variables bifurcation diagrams

# plot stable equilibria of  (Pstarset) against density of alternate prey (Yset)
plot(x = fset, y = Cstarset,
     type = 'l', lwd = 2, col = Ccol, las = 1,
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of coral cover, C*')

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = fset, y = Mstarset,
     type = 'l', lwd = 2, col = Mcol, las = 1, 
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of macroalgae cover, M*')

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = fset, y = Tstarset,
     type = 'l', lwd = 2, col = Tcol, las = 1, 
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of turf algae cover, T*')

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = fset, y = Pstarset,
     type = 'l', lwd = 2, col = Pcol, las = 1, 
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of parrotfish population, P*')
```
Variables (parrot fish population, macro algae cover, turf algae cover and coral cover) with respect to fishing pressure. As fishing pressure crosses a 0.38 threshold the population of parrot fish decreases to an eq of 0. Macro algae (the parrot fishes food choice) takes over the ground coverage and reaches an eq of 1, and the turf algae and coral cover reach an eq of 0.

Below is a closer look at the eq when fishing pressure hits 0.38.

```{r}
### Bifurcation diagram of fishing effort parameter vs. variables combined 

# sequence of values for f (fishing pressure) focused on the 0.38 threshold
fset <- seq(from = 0.35, to = 0.41,length.out = 50)

# holding vectors for equilibrium values of C, M, and T
Cstarset <- NaN*fset 
Mstarset <- NaN*fset
Tstarset <- NaN*fset
Pstarset <- NaN*fset

# for loop for fishing pressure bifurcation diagram
for(j in 1:length(fset)){
  # assign the value of Y
    f <- fset[j] 
    
    # create a holding vector for predator and prey populations
    # and fill with initial conditions
    C.simu_f <- NaN*tset
    C.simu_f[1] <- 1   
    M.simu_f <- NaN*tset 
    M.simu_f[1] <- 1
    T.simu_f <- NaN*tset 
    T.simu_f[1] <- 1
    P.simu_f <- NaN*tset 
    P.simu_f[1] <- 1
    
    for(i in 2:length(tset)){
      # calculating change in time
          # calculating change in time
      dt <- tset[i]-tset[i-1] 
    
      # set variable equal to the simulation value at the previous timestep
      M <- M.simu_f[i-1]
      C <- C.simu_f[i-1]
      Tt <- T.simu_f[i-1]
      P <- P.simu_f[i-1]
        
    # parrotfish population growth, assuming f = 0, and max carrying capacity = C
    dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
      # calculating change in population size -- replace g with P squiggle (P)
    dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
    #dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
    dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt

    
    # calculating total population size
    M.simu_f[i] <- M.simu_f[i-1] + dM
    C.simu_f[i] <- C.simu_f[i-1] + dC
    T.simu_f[i] <- 1 - C.simu_f[i] - M.simu_f[i]
      # P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to   coral cover.
    P.simu_f[i] <- P.simu_f[i-1] + dP
    } #end of the inner for loop
    
    # storing last population size (equilibrium population size) in holding vectors for equilibrium population sizes
    Cstarset[j] <- C.simu_f[length(tset)]
    Mstarset[j] <- M.simu_f[length(tset)]
    Tstarset[j] <- T.simu_f[length(tset)]
    Pstarset[j] <- P.simu_f[length(tset)]
    
} #end of outer for loop that iterates over each element in Yset
```


```{r}
# Bifurcation diagram with all variables
# plot stable equilibria

#macro aglae
plot(x = fset, y = Mstarset,
     type = 'l', lwd = 2, col = Mcol, las = 1,
     xlab = 'Fishing pressure (f)', ylab = 'Equilibria')

#coral
lines(x = fset, y = Cstarset,
     type = 'l', col = Ccol, lwd=2,)
#turf algae
lines(x = fset, y = Tstarset,
      col = Tcol, lwd = 2,)
#parrotfish
lines(x = fset, y = Pstarset,
     col = Pcol, lwd = 2, )

legend(x=0.385,y=0.68,
       legend=c("parrot fish population","macroalgae cover","turf algae cover","coral cover"),
       lty=1, lwd=3,
       col=c(Pcol,Mcol,Tcol,Ccol))
```


```{r}
####  Simulation 3: WITH PARROTFISH -- food availability as primary limiting resource ####

# K(C) = 1 - zC where 0 <= z < 1

## maybe we don't have to do this one, and can just focus on the first simulation??





```

