---
title: "Final Draft EEMB 179 Project Code"
author: "Sadie Cwikiel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recreating the Model from "The effect of fishing on hysteresis in Caribbean coral reefs” (Blackwood et al. 2012)

## Model Setup

### Equations

$$
\begin{align}
Macroalgae: \frac{d M}{dt}  &= \alpha M C - \frac{ gM}{(M + T)} + \gamma M T\\ 
\newline
Coral: \frac{d C}{dt} &= rTC - dC - \alpha M C\\
\newline
Turf Algae:  \frac{d T}{dt} &= \frac{ gM}{(M + T)} - \gamma M T + dC\\
\newline
Parrotfish: \frac{d P}{dt} &= s P(1 - \frac{P}{\beta K(C)}) - fP\\
\newline
\frac{d \tilde{P}}{dt} &= s \tilde{P}(1 - \frac{\tilde{P}}{K(C)}) - f\tilde{P}\\
\end{align}
$$

### Parameters
```{r}
# parrotfish growth rate
s <- 0.49 # intrinsic rate of growth s of parrotfish 

# rate macro algae directly overgrow coral per year
alpha <- 0.1 # coral overgrown by macro algae at a rate a

# rate macro algae spread vegetatively over algal turfs per year
gamma <- 0.8 # at a rate γ (gamma)

# rate of coral recruitment to algal turfs per year; coral always overgrow algal turfs
r <- 1 # corals recruit to and overgrow algal turfs at a rate r

# coral mortality -- natural coral mortality accounts for 2-4% per year, predation for 30%, and remaining mortality from coral disease or tissue damage following bleaching
d <- 0.44 # natural mortality rate of d

# grazing rate g can change, right now im setting it between the critical values 0.18 and 0.39
   # set grazing rate when it is above 0.39 coral will be in high equilibrium 
    #g <- 0.4
   # grazing rate when it is below 0.18 macroalgae will be in high equilibrium 
    #g <- 0.19
   # when grazing rate is between 0.18 and 0.39 and coral starts low, then bistability will bring back to low coral state, but when coral is high initially (above 0.5%), bistability will return to coral 
    #g <- 0.25

# maximum parrotfish carrying capacity
#beta <- C

# fishing pressure 
f <- 0

```

```{r}
## Store plot colors
Mcol <- 'limegreen' # plot color for macropalgae 
Tcol <- 'darkolivegreen4' # plot color for turf algae 
Ccol <- 'chocolate1' # plot color for coral 
Pcol <- 'darkturquoise' #plot color for parrotfish population
```

```{r}
#Initial Conditions:

## C and M start the same
#M0 <- .4 # macro algae starts at 40% concentration
#C0  <- .4 # coral starts at 40%
#T0  <- 1 - M0 - C0  # turf starts at 20% 

## Higher initial C
#M0 <- .2 # macro algae starts at 20% concentration
#C0  <- .65 # coral starts at 65%
#T0  <- 1 - M0 - C0  # turf starts at 15% 

## Highest initial C
#M0 <- .1 # macro algae starts at 10% concentration
#C0  <- .8 # coral starts at 80%
#T0  <- 1 - M0 - C0  # turf starts at 10% 

## High initial M
#M0 <- .8 # macro algae starts at 80% concentration
#C0  <- .1 # coral starts at 10%
#T0  <- 1 - M0 - C0  # turf starts at 10% 

```


##  Simulation 1: Original Model
constant grazing rate, no parrotfish population dynamics

initial conditions: mid g, low coral 
```{r}
# setting time and initial values 
# time set 0 to 100
tset <- seq(from = 0, to = 100, length.out = 20000) #set time from 0 to 100 

# set initial condition for the parameters
# M + C + T = 1 particular region of the seabed is covered entirely by macroalgae (M), coral (C), and algal turfs (T) (variables)
M0 <- .4 # macro algae starts at 40% concentration
C0  <- .4 # coral starts at 40%
T0  <- 1 - M0 - C0  # turf starts at 20% 

### Simulation 1: M, C, T dynamics in the absence of parrotfish grazing ###
g <- 0.25 #setting g to between threshold 

# creating holding vectors and filling initial conditions
M.simu <- NaN*tset 
M.simu[1] <- M0
C.simu <- NaN*tset
C.simu[1] <- C0
T.simu <- NaN*tset
T.simu[1] <- T0

#for loop for M, C, T dynamics in absence of parrotfish grazing
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu[i-1]
	C <- C.simu[i-1]
	Tt <- T.simu[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simu[i] <- M.simu[i-1] + dM
	C.simu[i] <- C.simu[i-1] + dC
	T.simu[i] <- 1 - C.simu[i] - M.simu[i]
}

# plot for M/C/T dynamics in the absence of parrotfish grazing
    # g = 0.4, C0 = 0.3
# plot macroalgae
plot(x = tset, y = M.simu,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu,
      lwd = 2, col = Ccol)

title("Original Model: M, C, and T Dynamics w/o dP/dt \n Start with LOW Coral Cover, g = 0.25")

legend(x = 1, y = .999,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```


initial conditions: mid g, high coral 
```{r}
# setting time and initial values 
# time set 0 to 100
tset <- seq(from = 0, to = 100, length.out = 20000) #set time from 0 to 100 

# set initial condition for the parameters, I just set these randomly but also to be equal to 1 as the paper says 
# M + C + T = 1 particular region of the seabed is covered entirely by macroalgae (M), coral (C), and algal turfs (T) (variables)
M0 <- .2 # macro algae starts at 20% concentration
C0  <- .65 # coral starts at 65%
T0  <- 1 - M0 - C0  # turf starts at 15% 

### Simulation 1: M, C, T dynamics in the absence of parrotfish grazing ###
g <- 0.25 #setting g to between threshold  

# creating holding vectors and filling initial conditions
M.simu1 <- NaN*tset 
M.simu1[1] <- M0
C.simu1 <- NaN*tset
C.simu1[1] <- C0
T.simu1 <- NaN*tset
T.simu1[1] <- T0

#for loop for M, C, T dynamics in absence of parrotfish grazing
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu1[i-1]
	C <- C.simu1[i-1]
	Tt <- T.simu1[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simu1[i] <- M.simu1[i-1] + dM
	C.simu1[i] <- C.simu1[i-1] + dC
	T.simu1[i] <- 1 - C.simu1[i] - M.simu1[i]
}

# plot for M/C/T dynamics in the absence of parrotfish grazing, HIGH coral cover 
# plot macroalgae
plot(x = tset, y = M.simu1,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu1,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu1,
      lwd = 2, col = Ccol)

title("Original Model: M, C, and T Dynamics w/o dP/dt \n Start with HIGH Coral Cover, g = 0.25")

legend(x = 60, y = 1,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```


initial conditions: low coral, g above threshold 
```{r}
# setting time and initial values 
# time set 0 to 100
tset <- seq(from = 0, to = 100, length.out = 20000) #set time from 0 to 100 

# set initial condition for the parameters, I just set these randomly but also to be equal to 1 as the paper says 
# M + C + T = 1 particular region of the seabed is covered entirely by macroalgae (M), coral (C), and algal turfs (T) (variables)
M0 <- .8 # macro algae starts at 80% concentration
C0  <- .1 # coral starts at 10%
T0  <- 1 - M0 - C0  # turf starts at 10% 

### Simulation 1: M, C, T dynamics in the absence of parrotfish grazing ###
g <- 0.4 #setting g to above threshold 

# creating holding vectors and filling initial conditions
M.simuG <- NaN*tset 
M.simuG[1] <- M0
C.simuG <- NaN*tset
C.simuG[1] <- C0
T.simuG <- NaN*tset
T.simuG[1] <- T0

#for loop for M, C, T dynamics in absence of parrotfish grazing
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simuG[i-1]
	C <- C.simuG[i-1]
	Tt <- T.simuG[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simuG[i] <- M.simuG[i-1] + dM
	C.simuG[i] <- C.simuG[i-1] + dC
	T.simuG[i] <- 1 - C.simuG[i] - M.simuG[i]
}

# plot for M/C/T dynamics in the absence of parrotfish grazing, low coral cover, high G 
# plot macroalgae
plot(x = tset, y = M.simuG,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simuG,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simuG,
      lwd = 2, col = Ccol)

title("Original Model: M, C, and T Dynamics w/o dP/dt \n Start with LOW Coral Cover, g = 0.4")

legend(x = 60, y = 1,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```


initial conditions: high coral, g below threshold 
```{r}
# setting time and initial values 
# time set 0 to 100
tset <- seq(from = 0, to = 100, length.out = 20000) #set time from 0 to 100 

# set initial condition for the parameters, I just set these randomly but also to be equal to 1 as the paper says 
# M + C + T = 1 particular region of the seabed is covered entirely by macroalgae (M), coral (C), and algal turfs (T) (variables)
M0 <- .1 # macro algae starts at 10% concentration
C0  <- .8 # coral starts at 80%
T0  <- 1 - M0 - C0  # turf starts at 10% 

### Simulation 1: M, C, T dynamics in the absence of parrotfish grazing ###
g <- 0.17 #setting g to above threshold 

# creating holding vectors and filling initial conditions
M.simuG <- NaN*tset 
M.simuG[1] <- M0
C.simuG <- NaN*tset
C.simuG[1] <- C0
T.simuG <- NaN*tset
T.simuG[1] <- T0

#for loop for M, C, T dynamics in absence of parrotfish grazing
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simuG[i-1]
	C <- C.simuG[i-1]
	Tt <- T.simuG[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simuG[i] <- M.simuG[i-1] + dM
	C.simuG[i] <- C.simuG[i-1] + dC
	T.simuG[i] <- 1 - C.simuG[i] - M.simuG[i]
}

# plot for M/C/T dynamics in the absence of parrotfish grazing, low coral cover, high G 
# plot macroalgae
plot(x = tset, y = M.simuG,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simuG,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simuG,
      lwd = 2, col = Ccol)

title("Original Model: M, C, and T Dynamics w/o dP/dt \n Start with LOW Coral Cover, g = 0.4")

legend(x = 60, y = 1,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```


### Bifurcation Diagram of Hysteresis
- explaining how Coral cover (C) depends on grazing rate (g)
- region of bistability
- phase shift and alternate stable states exhibiting hysteresis

```{r}
alpha <- 0.1 # coral overgrown by macroalgae at a rate a

# rate macroalgae spread vegetatively over algal turfs per year
gamma <- 0.8 # at a rate γ (gamma)

# rate of coral recruitment to algal turfs per year; coral always overgrow algal turfs
r <- 1 # corals recruit to and overgrow algal turfs at a rate r

# coral mortality -- natural coral mortality accounts for 2-4% per year, predation for 30%, and remaining mortality from coral disease or tissue damage following bleaching
d <- 0.44 # natural mortality rate of d



# setting a g set vector 
g_set <- seq(from = 0, to = 0.55, length.out = 100)


#set for C when it starts low: for region of bistability 
Mset <- NaN*g_set # holding vector for the equilibrium resource population size for each value of Y in Yset
Csetlow <- NaN*g_set # holding vector for N population size for each value of Y in Yset
Tset <- NaN*g_set # holding vector for M population size for each value of Y in Yset


#nested for loop  
for(j in 1:length(g_set)){ # for each element j from 1 to the length of Kset
	g <- g_set[j] # set g equal to the jth element of Kset
	
	# now simulate the model with this vector of Kset:
	# create holding vectors 
  M.simu <- NaN*tset; M.simu[1] <- 0.9
  C.simu <- NaN*tset; C.simu[1] <- 0.05
  T.simu <- NaN*tset; T.simu[1] <- 0.05

for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu[i-1]
	C <- C.simu[i-1]
	Tt <- T.simu[i-1]
		
	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simu[i] <- M.simu[i-1] + dM
	C.simu[i] <- C.simu[i-1] + dC
	T.simu[i] <- 1 - C.simu[i] - M.simu[i]
}
	
	# storing population sizes at the last timepoint of our simulation (the equilibrium population sizes) in our holding vectors for equilibrium population sizes 
  Mset[j] <- M.simu[length(tset)] 
	Csetlow[j] <- C.simu[length(tset)] 
	Tset[j] <- T.simu[length(tset)] 
	
}
```

```{r}
#set for C when it starts HIGH: for region of bistability 
Mset <- NaN*g_set # holding vector for the equilibrium resource population size for each value of Y in Yset
Csethigh <- NaN*g_set # holding vector for N population size for each value of Y in Yset
Tset <- NaN*g_set # holding vector for M population size for each value of Y in Yset


#nested for loop  
for(j in 1:length(g_set)){ # for each element j from 1 to the length of Kset
	g <- g_set[j] # set g equal to the jth element of Kset
	
	# now simulate the model with this vector of Kset:
	# create holding vectors 
  M.simu <- NaN*tset; M.simu[1] <- 0.05
  C.simu <- NaN*tset; C.simu[1] <- 0.9
  T.simu <- NaN*tset; T.simu[1] <- 0.05

for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu[i-1]
	C <- C.simu[i-1]
	Tt <- T.simu[i-1]
		
	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((g*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	
	# calculating total population size
	M.simu[i] <- M.simu[i-1] + dM
	C.simu[i] <- C.simu[i-1] + dC
	T.simu[i] <- 1 - C.simu[i] - M.simu[i]
}
	
	# storing population sizes at the last timepoint of our simulation (the equilibrium population sizes) in our holding vectors for equilibrium population sizes 
  Mset[j] <- M.simu[length(tset)] 
	Csethigh[j] <- C.simu[length(tset)] 
	Tset[j] <- T.simu[length(tset)] 
	
}
```


#### Coral Cover vs. Grazing Rate  

```{r}
# plot grazing as function of Coral cover

plot(x = g_set, y = Csetlow, #plot the grazing bifurcation when coral starts low
     type = 'l', las = 1, lwd = 2, col = '#FFBE9F', 
     xlab = 'Grazing intensity (g)', ylab = 'Equilibrium Coral Cover (C)',
     main = 'Coral Cover vs. Grazing intensity')
lines(x = g_set, y = Csethigh, col = '#DE4800', lwd = 2) #plot the grazing bifurcation when coral starts high

#showing bistability  
rect(xleft = 0.18, ybottom = -1, xright = 0.39, ytop = 0.7, col = rgb(0, 0, 0, alpha = 0.05), lty = 2, border = 1) #ligth grey box to show recovery
text(x = 0.29, y = 0.29, 'Region of Bistability', cex = 0.9) # text to signify bistability
text(x = 0.29, y = 0.54, 'start with high coral cover', cex = 0.6, col = '#DE4800') # text to signify low coral cover 
text(x = 0.29, y = 0.02, 'start with low coral cover', cex = 0.6, col = '#FFBE9F') # text to signify high coral cover 

text(x = 0.13, y = 0.54, 'Gcrit1 = 0.18', cex = 0.9) # text for g crit 1
text(x = 0.44, y = 0.02, 'Gcrit2 = 0.39', cex = 0.9) # text for g crit 2

```


## Simulation 2: Benthic cover dynamics influenced by parrotfish population dynamics (habitat as primary limiting resource of parrotfish popualtion)

Assumptions:
- Parrotfish populations grow logistically and influence the grazing rate, therefore impacting M, C, and T dynamics.
- Parrotfish rely on coral for habitat, and parrotfish population growth in simulation 2 is limited by coral cover.
- First, we assume that fishing pressure (f) = 0.
- The grazing rate of parrotfish is described as g(P) = (alpha*P)/beta. Alpha is a positive constant and is equal to the maximum grazing rate, so is assumed to be 1. Therefore, g(P) = P/beta, which guarantees that the grazing intensity arrives at a maximum only if f = 0 and P(t) reaches maximum carrying capacity (beta), requiring no limitation from habitat (i.d., K(C) = 1).
- For this simulation, we assume that beta = K(C) = C. 
- P squiggle is the equation of parrotfish population growth with respect to beta (max carrying capacity), and we use P to signify P squiggle in the simulation below. 
- P (squiggle) has been substituted in for g in the macroalgae cover equation because g(P) = Psquiggle. 

For these time series plots, we only include dM/dt, dC/dt, and dT/dt lines because the dP/dt line is parrotfish population relative to maximum carrying capacity, which in this model is just C so the dP/dt line ends up being the same as dC/dt. 


```{r}
####  Simulation 2: WITH PARROTFISH -- habitat as primary limiting resource ####

## Notes and Assumptions ##
# parrotfish logistic growth changes the grazing rate (g = P, see model assumptions) and therefore impacts M, C, and T dynamics
# assume that f = 0 (no fishing)
# g(P) = alpha*P/beta where alpha is a positive constant. # alpha = g_max, assumed to be 1, so g(P) = P/beta, which guarantees that the grazing intensity arrives at a maximum only if f = 0 and P(t) reaches maximum carrying capacity, requiring no limitation from habitat (i.e., K(C) = 1)
# so we can assume for this simulation that beta = K(C) = C
# P = P squiggle here, which is population size relative to maximum carrying capacity (beta)
# g(P) = P squiggle = P, so in the equation for dM, we've replaced g with P


## equal M and C initial conditions ##
M0 <- .4 # macro algae starts at 50% concentration
C0  <- .4 # coral starts at 30%
T0  <- 1 - M0 - C0  # turf starts at 20% 
P0 <- 0.5

# creating holding vectors and filling initial conditions
M.simu2 <- NaN*tset 
M.simu2[1] <- M0
T.simu2 <- NaN*tset
T.simu2[1] <- T0
C.simu2 <- NaN*tset
C.simu2[1] <- C0
P.simu2 <- NaN*tset
P.simu2[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu2[i-1]
	C <- C.simu2[i-1]
	Tt <- T.simu2[i-1]
	P <- P.simu2[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	
	# calculating total population size
	M.simu2[i] <- M.simu2[i-1] + dM
	C.simu2[i] <- C.simu2[i-1] + dC
	T.simu2[i] <- 1 - C.simu2[i] - M.simu2[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu2[i] <- P.simu2[i-1] + dP
}


## PLOT

# plot macroalgae
plot(x = tset, y = M.simu2,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu2,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu2,
      lwd = 2, col = Ccol)
#plot parrotfish
# lines(x = tset, y = P.simu2,
#       lwd = 2, col = Pcol)

title("M, C, and T Dynamics with dP/dt \n initial C = initial M")

legend(x = 60, y = 0.95,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))

```
Parrotfish abundance relative to maximum carrying capacity (beta) over time. The population declines rapidly at the beginning of the simulation when coral cover is low because the proportion of coral cover is assumed to be the same as the carrying capacity in this habitat-limited scenario. As coral cover increases and levels off throughout the simulation, the parrotfish population does the same. 


```{r}
## Simulation 2: higher initial C

## Higher initial C
M0 <- .2 # macro algae starts at 20% concentration
C0  <- .65 # coral starts at 65%
T0  <- 1 - M0 - C0  # turf starts at 15% 
P0 <- 0.5

# creating holding vectors and filling initial conditions
M.simu3 <- NaN*tset 
M.simu3[1] <- M0
T.simu3 <- NaN*tset
T.simu3[1] <- T0
C.simu3 <- NaN*tset
C.simu3[1] <- C0
P.simu3 <- NaN*tset
P.simu3[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu3[i-1]
	C <- C.simu3[i-1]
	Tt <- T.simu3[i-1]
	P <- P.simu3[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	
	# calculating total population size
	M.simu3[i] <- M.simu3[i-1] + dM
	C.simu3[i] <- C.simu3[i-1] + dC
	T.simu3[i] <- 1 - C.simu3[i] - M.simu3[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu3[i] <- P.simu3[i-1] + dP
}


## PLOT

# plot macroalgae
plot(x = tset, y = M.simu3,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu3,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu3,
      lwd = 2, col = Ccol)
#plot parrotfish
# lines(x = tset, y = P.simu3,
#       lwd = 2, col = Pcol, lty = 2)

title("M, C, and T Dynamics with dP/dt \n high initial C, low initial M")

legend(x = 60, y = 0.95,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```

```{r}
## Simulation 2: highest initial C

## Highest initial C
M0 <- .1 # macro algae starts at 10% concentration
C0  <- .8 # coral starts at 80%
T0  <- 1 - M0 - C0  # turf starts at 10% 
P0 <- 0.5

# creating holding vectors and filling initial conditions
M.simu4 <- NaN*tset 
M.simu4[1] <- M0
T.simu4 <- NaN*tset
T.simu4[1] <- T0
C.simu4 <- NaN*tset
C.simu4[1] <- C0
P.simu4 <- NaN*tset
P.simu4[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu4[i-1]
	C <- C.simu4[i-1]
	Tt <- T.simu4[i-1]
	P <- P.simu4[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	
	# calculating total population size
	M.simu4[i] <- M.simu4[i-1] + dM
	C.simu4[i] <- C.simu4[i-1] + dC
	T.simu4[i] <- 1 - C.simu4[i] - M.simu4[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu4[i] <- P.simu4[i-1] + dP
}


## PLOT

# plot macroalgae
plot(x = tset, y = M.simu4,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu4,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu4,
      lwd = 2, col = Ccol)
#plot parrotfish
# lines(x = tset, y = P.simu3,
#       lwd = 2, col = Pcol, lty = 2)

title("M, C, and T Dynamics with dP/dt \n very high initial C, very low initial M")

legend(x = 60, y = 0.95,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```


```{r}
## Simulation 2: highest initial M, lowest C

## High initial M, low C
M0 <- .8 # macro algae starts at 80% concentration
C0  <- .1 # coral starts at 10%
T0  <- 1 - M0 - C0  # turf starts at 10% 
P0 <- 0.5

# creating holding vectors and filling initial conditions
M.simu5 <- NaN*tset 
M.simu5[1] <- M0
T.simu5 <- NaN*tset
T.simu5[1] <- T0
C.simu5 <- NaN*tset
C.simu5[1] <- C0
P.simu5 <- NaN*tset
P.simu5[1] <- P0

#for loop for the simulation
for(i in 2:length(tset)){
  # calculating change in time
	dt <- tset[i]-tset[i-1] 
	
	# set variable equal to the simulation value at the previous timestep
	M <- M.simu5[i-1]
	C <- C.simu5[i-1]
	Tt <- T.simu5[i-1]
	P <- P.simu5[i-1]
		

	# calculating change in population size -- replace g with P squiggle (P)
	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt
	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	
	# calculating total population size
	M.simu5[i] <- M.simu5[i-1] + dM
	C.simu5[i] <- C.simu5[i-1] + dC
	T.simu5[i] <- 1 - C.simu5[i] - M.simu5[i]
	# P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to coral cover.
	P.simu5[i] <- P.simu5[i-1] + dP
}


## PLOT

# plot macroalgae
plot(x = tset, y = M.simu5,
     type = 'l', las = 1, lwd = 2, col = Mcol,
     xlab = 'Time', ylab = 'Proportion of Coral, Turf and Macroalgae',
     xlim = c(min(tset), max(tset)), ylim = c(0, 1))
# plot turf
lines(x = tset, y = T.simu5,
      lwd = 2, col = Tcol)
# plot coral
lines(x = tset, y = C.simu5,
      lwd = 2, col = Ccol)
#plot parrotfish
# lines(x = tset, y = P.simu3,
#       lwd = 2, col = Pcol, lty = 2)

title("M, C, and T Dynamics with dP/dt \n very low initial C, very high initial M")

legend(x = 60, y = 0.95,
       c('Macroalgae', 'Turf Algae', 'Coral'),
       lwd = 2,
       col = c(Mcol, Tcol, Ccol))
```

Here, because the carrying capacity of parrotfish scales with coral cover, the system cannot shift from a low coral state to a high coral state.


## Bifurcation Diagrams of state variables vs. fishing pressure
```{r}
### Bifurcation diagram of fishing effort parameter vs. coral cover equilibrium value

# sequence of values for f (fishing pressure)
fset <- seq(from = 0, to = 1,length.out = 50)

# holding vectors for equilibrium values of C, M, and T
Cstarset <- NaN*fset 
Mstarset <- NaN*fset
Tstarset <- NaN*fset
Pstarset <- NaN*fset

# for loop for fishing pressure bifurcation diagram
for(j in 1:length(fset)){
  # assign the value of Y
	f <- fset[j] 
	
	# create a holding vector for predator and prey populations
	# and fill with initial conditions
	C.simu_f <- NaN*tset
	C.simu_f[1] <- 1   
	M.simu_f <- NaN*tset 
	M.simu_f[1] <- 1
	T.simu_f <- NaN*tset 
	T.simu_f[1] <- 1
	P.simu_f <- NaN*tset 
	P.simu_f[1] <- 1
	
	for(i in 2:length(tset)){
	  # calculating change in time
		  # calculating change in time
	  dt <- tset[i]-tset[i-1] 
	
	  # set variable equal to the simulation value at the previous timestep
	  M <- M.simu_f[i-1]
	  C <- C.simu_f[i-1]
	  Tt <- T.simu_f[i-1]
	  P <- P.simu_f[i-1]
		
  	# parrotfish population growth, assuming f = 0, and max carrying capacity = C
  	dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
	  # calculating change in population size -- replace g with P squiggle (P)
  	dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
  	#dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
  	dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt

	
  	# calculating total population size
  	M.simu_f[i] <- M.simu_f[i-1] + dM
  	C.simu_f[i] <- C.simu_f[i-1] + dC
  	T.simu_f[i] <- 1 - C.simu_f[i] - M.simu_f[i]
	  # P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to   coral cover.
  	P.simu_f[i] <- P.simu_f[i-1] + dP
  	} #end of the inner for loop
	
	# storing last population size (equilibrium population size) in holding vectors for equilibrium population sizes
	Cstarset[j] <- C.simu_f[length(tset)]
	Mstarset[j] <- M.simu_f[length(tset)]
	Tstarset[j] <- T.simu_f[length(tset)]
	Pstarset[j] <- P.simu_f[length(tset)]
	
} #end of outer for loop that iterates over each element in Yset

```

```{r}
## Plot for fishing pressure vs. variables bifurcation diagrams

# plot stable equilibria of  (Pstarset) against density of alternate prey (Yset)
plot(x = fset, y = Cstarset,
     type = 'l', lwd = 2, col = Ccol, las = 1,
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of coral cover, C*')

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = fset, y = Mstarset,
     type = 'l', lwd = 2, col = Mcol, las = 1, 
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of macroalgae cover, M*')

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = fset, y = Tstarset,
     type = 'l', lwd = 2, col = Tcol, las = 1, 
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of turf algae cover, T*')

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = fset, y = Pstarset,
     type = 'l', lwd = 2, col = Pcol, las = 1, 
     xlab = 'Fishing pressure (f)', ylab = 'Stable equilibria of parrotfish population, P*')
```


Variables (parrot fish population, macro algae cover, turf algae cover and coral cover) with respect to fishing pressure. As fishing pressure crosses a 0.38 threshold the population of parrot fish decreases to an eq of 0. Macro algae (the parrot fishes food choice) takes over the ground coverage and reaches an eq of 1, and the turf algae and coral cover reach an eq of 0.

Below is a closer look at the eq when fishing pressure hits 0.38.

```{r}
### Bifurcation diagram of fishing effort parameter vs. variables combined 

# sequence of values for f (fishing pressure) focused on the 0.38 threshold
fset <- seq(from = 0.35, to = 0.41,length.out = 50)

# holding vectors for equilibrium values of C, M, and T
Cstarset <- NaN*fset 
Mstarset <- NaN*fset
Tstarset <- NaN*fset
Pstarset <- NaN*fset

# for loop for fishing pressure bifurcation diagram
for(j in 1:length(fset)){
  # assign the value of Y
    f <- fset[j] 
    
    # create a holding vector for predator and prey populations
    # and fill with initial conditions
    C.simu_f <- NaN*tset
    C.simu_f[1] <- 1   
    M.simu_f <- NaN*tset 
    M.simu_f[1] <- 1
    T.simu_f <- NaN*tset 
    T.simu_f[1] <- 1
    P.simu_f <- NaN*tset 
    P.simu_f[1] <- 1
    
    for(i in 2:length(tset)){
      # calculating change in time
          # calculating change in time
      dt <- tset[i]-tset[i-1] 
    
      # set variable equal to the simulation value at the previous timestep
      M <- M.simu_f[i-1]
      C <- C.simu_f[i-1]
      Tt <- T.simu_f[i-1]
      P <- P.simu_f[i-1]
        
    # parrotfish population growth, assuming f = 0, and max carrying capacity = C
    dP <- ((s*P)*(1 - (P/C)) - f*P) * dt
      # calculating change in population size -- replace g with P squiggle (P)
    dM <- ( (alpha*M*C) - ((P*M)/(M + Tt)) + (gamma*M*Tt) ) * dt 
    #dT <- ( (g*M)/(M + T) - (gamma*M*T) - (r*T*C) + (d*C)) * dt 
    dC <- ((r*Tt*C) - (d*C) - (alpha*M*C) ) * dt

    
    # calculating total population size
    M.simu_f[i] <- M.simu_f[i-1] + dM
    C.simu_f[i] <- C.simu_f[i-1] + dC
    T.simu_f[i] <- 1 - C.simu_f[i] - M.simu_f[i]
      # P here is population relative to maximum carrying capacity (beta), which also equals coral cover. So P = population relative to   coral cover.
    P.simu_f[i] <- P.simu_f[i-1] + dP
    } #end of the inner for loop
    
    # storing last population size (equilibrium population size) in holding vectors for equilibrium population sizes
    Cstarset[j] <- C.simu_f[length(tset)]
    Mstarset[j] <- M.simu_f[length(tset)]
    Tstarset[j] <- T.simu_f[length(tset)]
    Pstarset[j] <- P.simu_f[length(tset)]
    
} #end of outer for loop that iterates over each element in Yset
```


```{r}
# Bifurcation diagram with all variables
# plot stable equilibria

#macro aglae
plot(x = fset, y = Mstarset,
     type = 'l', lwd = 2, col = Mcol, las = 1,
     xlab = 'Fishing pressure (f)', ylab = 'Equilibria')

#coral
lines(x = fset, y = Cstarset,
     type = 'l', col = Ccol, lwd=2,)
#turf algae
lines(x = fset, y = Tstarset,
      col = Tcol, lwd = 2,)
#parrotfish
lines(x = fset, y = Pstarset,
     col = Pcol, lwd = 2, )

title("Bifurcation Diagram for Fishing Pressure")

legend(x=0.388,y=0.68,
       legend=c("parrotfish population","macroalgae cover","turf algae cover","coral cover"),
       lty=1, lwd=3,
       col=c(Pcol,Mcol,Tcol,Ccol))
```






